---
title: "Homework 7"
author: "Hannah Damico"
date: "2023-02-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
```

## Read in Data

```{r}
epese_data <- read.csv("./epese_bmi2-1.csv")
```


## Check out Data & Make some changes

```{r}
head (epese_data)
summary(epese_data[, 2:25]) # summary statistics on age time homework 2 data set.


# reverse code dead to deal with censor since death is the event, the censor would be alive
# Now censor = 1 is alive, censor = 0 is dead
epese_data$censor = -1 * (epese_data$dead - 1)
```


# Look at the  survival by  smoking group

```{r}
surv_probs <-
  survfit(Surv(age_at_death, dead) ~ as.factor(smoke_at_baseline), data = epese_data)
# summary(surv_probs)
```


# check the log rank statistic

```{r}

# Tests if there is a difference between two or more survival curves using the G^\rho
# family of tests, or for a single curve against a known alternative

survdiff(Surv(age_at_death, dead) ~ as.factor(smoke_at_baseline), data = epese_data)
```


# plot

```{r}

plot (
  surv_probs,
  main = "K-M estimates for smoking baseline, EPESE",
  conf.int = FALSE,
  xlab = "years",
  ylab = "survival function"
)
summary(surv_probs)
```


# Question 1: Does smoking matter PHreg?

According to the KM plot above, there are differences in survival across smoking groups. We make this conclusion since the KM curves for smokers & non-smokers do not cross, indicating a difference in survival across groups.

```{r}

fit1.survival <-
  coxph(
    Surv(age_at_death, dead) ~ smoke_at_baseline,
    method = "breslow",
    na.action = na.exclude,
    data = epese_data
  )
summary (fit1.survival)

plot(survfit(fit1.survival, fun = "cloglog"))
```



```{r}
library(survival)
data(lung)
fit <- coxph(Surv(time, status) ~ sex, data = lung)
plot(survfit(fit, fun = "cloglog"), mark.time = FALSE)

```



# Question 3: race effect


```{r}
fit3.survival <- coxph(
  Surv(age_at_death, dead) ~ white,
  method = "breslow",
  na.action = na.exclude,
  data = epese_data
)
summary (fit3.survival)
```

# plot survival curves

We see an intersection of HR curves across groups, indicating that at some point in time (t = 15?), the HR are equivalent. 

```{r}
surv_probs <-
  survfit(Surv(age_at_death, dead) ~ as.factor(white), data = epese_data)
surv_probs2 <-
  (surv_probs)    #Can also use summary(surv_probs). Removes censored values but also removes final values
prob <- as.data.frame(cbind(surv_probs2$surv, surv_probs2$time))
names(prob) <- c('surv', 'time')
prob$white = NA
prob

prob$white[1:36] = 0
prob$white[37:70] = 1

plot (
  prob$time[prob$white == 1],
  prob$surv[prob$white == 1],
  type = 'p',
  col = c(4),
  xlim = c(0, 38),
  ylim = c(0, 1),
  main = " survival func: race",
  xlab = "age at death -70",
  ylab = "survival probability"
)
par(new = T)
plot (
  prob$time[prob$white == 0],
  prob$surv[prob$white == 0],
  type = 'p',
  col = c(2),
  xlim = c(0, 38),
  ylim = c(0, 1),
  ylab = '',
  xlab = ''
)
lines(
  prob$time[prob$white == 1],
  prob$surv[prob$white == 1],
  type = 'l',
  col = c(4),
  xlim = c(0, 38),
  ylim = c(0, 1),
  ylab = '',
  xlab = ''
)
lines(
  prob$time[prob$white == 0],
  prob$surv[prob$white == 0],
  type = 'l',
  col = c(2),
  xlim = c(0, 38),
  ylim = c(0, 1),
  ylab = '',
  xlab = ''
)

```



```{r}
surv_probs <-
  survfit(Surv(age_at_death, dead) ~ as.factor(white), data = epese_data)
surv_probs2 <-
  (surv_probs)    #Can also use summary(surv_probs). Removes censored values but also removes final values
surv_probs2$lls_surv <- log(-log(surv_probs2$surv))
surv_probs2$time <- log(surv_probs2$time)
log_prob <-
  as.data.frame(cbind(surv_probs2$lls_surv, surv_probs2$time))
names(log_prob) <- c('surv', 'time')
log_prob$white = NA
x <- max(as.numeric(rownames(log_prob[log_prob$time == 0, ])))
log_prob$white[1:x - 1] = 1
log_prob$white[x:nrow(log_prob)] = 0

plot (
  log_prob$time[log_prob$white == 1],
  log_prob$surv[log_prob$white == 1],
  type = 'p',
  col = c(4),
  xlim = c(0, 4),
  ylim = c(-6.5, 1),
  main = "Log of -log of estimated surv func: race",
  xlab = "log(age at death) -70",
  ylab = "log(-log(survival function))"
)
par(new = T)
plot (
  log_prob$time[log_prob$white == 0],
  log_prob$surv[log_prob$white == 0],
  type = 'p',
  col = c(2),
  xlim = c(0, 4),
  ylim = c(-6.5, 1),
  ylab = '',
  xlab = ''
)
lines(
  log_prob$time[log_prob$white == 1],
  log_prob$surv[log_prob$white == 1],
  type = 'l',
  col = c(4),
  xlim = c(0, 4),
  ylim = c(-6.5, 1),
  ylab = '',
  xlab = ''
)
lines(
  log_prob$time[log_prob$white == 0],
  log_prob$surv[log_prob$white == 0],
  type = 'l',
  col = c(2),
  xlim = c(0, 4),
  ylim = c(-6.5, 1),
  ylab = '',
  xlab = ''
)
```




